---
title: 'Chapter 5: R Objects'
author: "Kalpana Renjarla"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
  word_document:
    toc: true
subtitle: "Hands-On Programming with R — Sections 5.1, 5.3, 5.4, 5.7, and 5.8"
---

```{r setup, include=TRUE}
# Show all R code in the knitted document, as required by the assignment.
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE, comment = "##")
```

# Introduction

This documented R Markdown program replicates and executes every line of code in Sections 5.1, 5.3, 5.4, 5.7, and 5.8 of Garrett Grolemund's *Hands-On Programming with R*. Comments explain the purpose of each operation. The selected sections do not create graphs or charts, so graph interpretation is not applicable.

# Section 5.1 — Atomic Vectors

Atomic vectors are one-dimensional objects whose elements all share one data type. This section demonstrates R's six atomic-vector types: double, integer, character, logical, complex, and raw.

```{r atomic-vector-basics}
# Create a numeric atomic vector representing the six faces of a die.
die <- c(1, 2, 3, 4, 5, 6)

# Display the vector.
die

# Confirm that die is a vector.
is.vector(die)

# Create and display a one-element vector.
five <- 5
five

# Confirm its vector status and compare vector lengths.
is.vector(five)
length(five)
length(die)

# Create one-element integer and character vectors.
int <- 1L
text <- "ace"

# Create multi-element integer and character vectors.
int <- c(1L, 5L)
text <- c("ace", "hearts")

# Sum works with the integer vector.
sum(int)
```

The next line is intentionally reproduced from the book. It demonstrates that `sum()` cannot add character strings. The chunk permits the expected error so knitting can continue.

```{r expected-character-error, error=TRUE}
# Expected error: character strings cannot be summed.
sum(text)
```

## 5.1.1 Doubles

```{r doubles}
# Recreate the die; ordinary numeric literals are stored as doubles.
die <- c(1, 2, 3, 4, 5, 6)
die

# Identify the atomic type.
typeof(die)
```

## 5.1.2 Integers

```{r integers}
# The L suffix explicitly creates integer values.
int <- c(-1L, 2L, 4L)
int
typeof(int)

# Demonstrate a small floating-point rounding error with doubles.
sqrt(2)^2 - 2
```

## 5.1.3 Characters

```{r characters}
# Create and inspect a character vector.
text <- c("Hello", "World")
text
typeof(text)

# A single quoted word is also a character vector of length one.
typeof("Hello")
```

### Exercise 5.1 — Character or Number?

**Question:** Which of `1`, `"1"`, and `"one"` are character strings, and which are numbers?

**Answer:** `1` is numeric. `"1"` and `"one"` are character strings because quotation marks cause R to store their contents as character data.

```{r exercise-5-1}
# Verify the answer by inspecting each value's type.
typeof(1)
typeof("1")
typeof("one")
```

## 5.1.4 Logicals

```{r logicals}
# Evaluate a comparison; 3 is not greater than 4.
3 > 4

# Create and inspect a logical vector.
logic <- c(TRUE, FALSE, TRUE)
logic
typeof(logic)

# F is R's shorthand for FALSE unless F has been reassigned.
typeof(F)
```

## 5.1.5 Complex and Raw

```{r complex-and-raw}
# Create and inspect complex numbers containing imaginary components.
comp <- c(1 + 1i, 1 + 2i, 1 + 3i)
comp
typeof(comp)

# Create three empty raw bytes and inspect their type.
raw(3)
typeof(raw(3))
```

### Exercise 5.2 — Vector of Cards

**Question:** Create an atomic vector that stores the face names of the five cards in a royal flush. Which vector type is appropriate?

**Answer:** A character vector is appropriate because card faces are text labels.

```{r exercise-5-2}
# Store the five royal-flush face names in a character vector.
hand <- c("ace", "king", "queen", "jack", "ten")
hand

# Confirm that the object stores character data.
typeof(hand)
```

# Section 5.3 — Matrices

Matrices arrange one atomic vector into two dimensions. Because matrices are built from atomic vectors, all cells retain the same underlying data type.

```{r matrices}
# Arrange die into two rows; R fills columns by default.
m <- matrix(die, nrow = 2)
m

# Arrange the same values by rows instead of columns.
m <- matrix(die, nrow = 2, byrow = TRUE)
m
```

### Exercise 5.3 — Make a Matrix

**Question:** Create a five-row, two-column matrix containing the face and suit of every card in a spade royal flush.

**Answer:** The book demonstrates several equivalent constructions.

```{r exercise-5-3}
# List all faces first and all suits second.
hand1 <- c("ace", "king", "queen", "jack", "ten", "spades", "spades",
  "spades", "spades", "spades")

# Method 1: specify five rows.
matrix(hand1, nrow = 5)

# Method 2: specify two columns.
matrix(hand1, ncol = 2)

# Method 3: directly assign the dimensions attribute.
dim(hand1) <- c(5, 2)
hand1

# Alternatively, list each face beside its suit.
hand2 <- c("ace", "spades", "king", "spades", "queen", "spades", "jack",
  "spades", "ten", "spades")

# Fill row by row while specifying either rows or columns.
matrix(hand2, nrow = 5, byrow = TRUE)
matrix(hand2, ncol = 2, byrow = TRUE)
```

# Section 5.4 — Arrays

Arrays extend matrices to three or more dimensions. The example creates three 2-by-2 slices.

```{r arrays}
# Combine three number ranges and arrange them into a 2 x 2 x 3 array.
ar <- array(c(11:14, 21:24, 31:34), dim = c(2, 2, 3))

# Display all three array slices.
ar
```

**Interpretation:** Slice 1 contains values 11–14, slice 2 contains 21–24, and slice 3 contains 31–34. R fills each slice by columns.

# Section 5.7 — Lists

Unlike an atomic vector, a list can group R objects of different types and lengths, including another list.

```{r lists}
# Create a list containing a numeric sequence, text, and a nested logical list.
list1 <- list(100:130, "R", list(TRUE, FALSE))

# Display the nested list structure.
list1
```

### Exercise 5.5 — Use a List to Make a Card

**Question:** Use a list to store the face, suit, and point value of the ace of hearts.

**Answer:** A list preserves the face and suit as character values while keeping the point value numeric.

```{r exercise-5-5}
# Store three differently typed properties in separate list elements.
card <- list("ace", "hearts", 1)
card
```

# Section 5.8 — Data Frames

A data frame is a two-dimensional list. Each column is a vector, columns can have different types, and all columns must have equal lengths.

```{r simple-data-frame}
# Build a three-row data frame with face, suit, and numeric value columns.
df <- data.frame(face = c("ace", "two", "six"),
  suit = c("clubs", "clubs", "clubs"), value = c(1, 2, 3))

# Display the data frame.
df

# Demonstrate named elements in a list and vector.
list(face = "ace", suit = "hearts", value = 1)
c(face = "ace", suit = "hearts", value = "one")

# Verify that a data frame is a list with class data.frame.
typeof(df)
class(df)

# Display the internal structure of each column.
str(df)
```

**Version note:** The book was written when `data.frame()` converted text columns to factors by default. Current R versions default to character columns, so `str(df)` may display `chr` instead of the factor output printed in the book. This is an expected version difference, not an error.

```{r strings-as-factors-false}
# Explicitly prevent conversion of character strings to factors.
df <- data.frame(face = c("ace", "two", "six"),
  suit = c("clubs", "clubs", "clubs"), value = c(1, 2, 3),
  stringsAsFactors = FALSE)

# Verify the resulting data frame and its column types.
df
str(df)
```

## Complete 52-Card Data Frame

```{r complete-deck}
# Construct the complete playing-card deck exactly as shown in Section 5.8.
deck <- data.frame(
  face = c("king", "queen", "jack", "ten", "nine", "eight", "seven", "six",
    "five", "four", "three", "two", "ace", "king", "queen", "jack", "ten",
    "nine", "eight", "seven", "six", "five", "four", "three", "two", "ace",
    "king", "queen", "jack", "ten", "nine", "eight", "seven", "six", "five",
    "four", "three", "two", "ace", "king", "queen", "jack", "ten", "nine",
    "eight", "seven", "six", "five", "four", "three", "two", "ace"),
  suit = c("spades", "spades", "spades", "spades", "spades", "spades",
    "spades", "spades", "spades", "spades", "spades", "spades", "spades",
    "clubs", "clubs", "clubs", "clubs", "clubs", "clubs", "clubs", "clubs",
    "clubs", "clubs", "clubs", "clubs", "clubs", "diamonds", "diamonds",
    "diamonds", "diamonds", "diamonds", "diamonds", "diamonds", "diamonds",
    "diamonds", "diamonds", "diamonds", "diamonds", "diamonds", "hearts",
    "hearts", "hearts", "hearts", "hearts", "hearts", "hearts", "hearts",
    "hearts", "hearts", "hearts", "hearts", "hearts"),
  value = c(13, 12, 11, 10, 9, 8, 7, 6, 5, 4, 3, 2, 1, 13, 12, 11, 10, 9, 8,
    7, 6, 5, 4, 3, 2, 1, 13, 12, 11, 10, 9, 8, 7, 6, 5, 4, 3, 2, 1, 13, 12, 11,
    10, 9, 8, 7, 6, 5, 4, 3, 2, 1)
)

# Verify the completed deck and its dimensions.
deck
dim(deck)
str(deck)
```

**Interpretation:** The result contains 52 observations (one card per row) and three variables: `face`, `suit`, and `value`. The face and suit columns contain text, while value contains numeric point values.

# Conclusion

The replicated code demonstrates how R builds increasingly flexible objects: atomic vectors store one data type, matrices and arrays organize atomic data across dimensions, lists combine objects of different types, and data frames organize equal-length vectors as analysis-ready columns. All requested code from Sections 5.1, 5.3, 5.4, 5.7, and 5.8 has been documented and included.

# Reference

Grolemund, G. (2014). *Hands-on programming with R*. O'Reilly Media. https://rstudio-education.github.io/hopr/r-objects.html
