---
title: "Section 1.2.5 Exercises 3, 4, and 6"
author: "Kalpana Renjarla"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=TRUE}
# Display all R code in the knitted HTML file.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE)

# Load the packages required for the exercises.
library(tidyverse)
library(palmerpenguins)
```

# Introduction

This assignment answers Exercises 3, 4, and 6 from Section 1.2.5 of
*R for Data Science* (2nd ed.). The exercises use the `penguins` dataset
from the `palmerpenguins` package.

# Exercise 3

**Question:** Make a scatterplot of `bill_depth_mm` versus `bill_length_mm`,
with `bill_depth_mm` on the y-axis and `bill_length_mm` on the x-axis.
Describe the relationship between the two variables.

```{r exercise-3, fig.width=8, fig.height=5.5}
# Create a scatterplot with bill length on the x-axis and bill depth on the
# y-axis. Color is mapped to species to make the clusters easier to interpret.
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm,
    y = bill_depth_mm,
    color = species
  )
) +
  geom_point(na.rm = TRUE) +
  labs(
    title = "Penguin Bill Depth versus Bill Length",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  theme_minimal()
```

**Interpretation:** The relationship is not represented well by one simple
straight line because the points form clusters associated with the three
penguin species. Gentoo penguins generally have longer but shallower bills,
while Chinstrap penguins tend to have longer and deeper bills. Adelie penguins
usually have shorter bills. Within the species clusters, bill depth and bill
length tend to have a positive relationship, but the combined dataset shows
that species is an important variable for understanding the pattern.

**What we did:** We used `ggplot()` to compare two numerical variables:
`bill_length_mm` on the x-axis and `bill_depth_mm` on the y-axis. We used
`geom_point()` to create the scatterplot and assigned a different color to each
species so that the groups could be compared more clearly. We also used
`na.rm = TRUE` to exclude rows with missing bill measurements.

**What we achieved:** We created a clear scatterplot that revealed three
species-related clusters. The result showed that the relationship between bill
length and bill depth changes among Adelie, Chinstrap, and Gentoo penguins.
Therefore, species helps explain the pattern better than one overall trend
line would.

# Exercise 4

**Question:** What happens if you make a scatterplot of `species` versus
`bill_depth_mm`? What might be a better choice of geom?

```{r exercise-4-scatterplot, fig.width=8, fig.height=5.5}
# A point geom places observations into three vertical groups because species
# is categorical while bill depth is numerical. Many points overlap.
ggplot(
  data = penguins,
  mapping = aes(x = species, y = bill_depth_mm)
) +
  geom_point(na.rm = TRUE) +
  labs(
    title = "Bill Depth by Penguin Species: Point Plot",
    x = "Species",
    y = "Bill depth (mm)"
  ) +
  theme_minimal()
```

The point plot creates three vertical groups, one for each species. Because
many penguins have similar bill-depth measurements, the points overlap and the
distribution within each species is difficult to compare.

```{r exercise-4-boxplot, fig.width=8, fig.height=5.5}
# A boxplot is a better choice for comparing the distribution of a numerical
# variable across the levels of a categorical variable.
ggplot(
  data = penguins,
  mapping = aes(x = species, y = bill_depth_mm, fill = species)
) +
  geom_boxplot(na.rm = TRUE, show.legend = FALSE) +
  labs(
    title = "Distribution of Bill Depth by Penguin Species",
    x = "Species",
    y = "Bill depth (mm)"
  ) +
  theme_minimal()
```

**Interpretation:** A boxplot is a better geom because it summarizes the
median, spread, and possible outliers for each species. The graph shows that
Gentoo penguins generally have shallower bills than Adelie and Chinstrap
penguins. Chinstrap penguins have the highest typical bill depth, while Adelie
penguins fall between the other two groups.

**What we did:** We first plotted the categorical variable `species` against
the numerical variable `bill_depth_mm` with `geom_point()`. Because many points
overlapped within three vertical groups, we created a second graph with
`geom_boxplot()` to summarize and compare the bill-depth distributions.

**What we achieved:** We identified the boxplot as the more appropriate geom
for this comparison. It made the median, spread, and possible outliers easier
to see and showed clear differences in typical bill depth among the three
penguin species.

# Exercise 6

**Question:** What does the `na.rm` argument do in `geom_point()`? What is its
default value? Create a scatterplot that successfully uses this argument set
to `TRUE`.

The `na.rm` argument controls how missing values are handled by the geom. Its
default value is `FALSE`. With the default, rows that contain missing values
cannot be plotted and ggplot2 removes them while printing a warning. When
`na.rm = TRUE`, those incomplete rows are removed silently, so the scatterplot
is created without the missing-value warning.

```{r exercise-6, fig.width=8, fig.height=5.5}
# Remove observations with missing bill-length or bill-depth values silently
# and create the requested scatterplot.
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) +
  geom_point(na.rm = TRUE, color = "#2C7FB8", alpha = 0.75) +
  labs(
    title = "Bill Measurements with Missing Values Removed",
    subtitle = "geom_point(na.rm = TRUE)",
    x = "Bill length (mm)",
    y = "Bill depth (mm)"
  ) +
  theme_minimal()
```

**Interpretation:** The scatterplot displays all observations that have both a
bill-length value and a bill-depth value. Rows missing either measurement are
excluded silently. Setting `na.rm = TRUE` changes the handling of the warning;
it does not estimate or replace the missing measurements.

**What we did:** We examined the purpose and default setting of the `na.rm`
argument in `geom_point()`. We then set `na.rm = TRUE` while creating a
scatterplot of bill length and bill depth so that observations with missing
values would be removed without producing the usual warning.

**What we achieved:** The graph was created successfully using all complete
bill-measurement records. We also demonstrated that `na.rm = TRUE` removes
incomplete rows silently but does not fill in, estimate, or otherwise change
the missing values.

# Conclusion

These exercises demonstrate that choosing an appropriate geom depends on the
types of variables being compared. A scatterplot is suitable for two numerical
variables, while a boxplot is more informative for comparing a numerical
variable across categories. They also show how `na.rm = TRUE` allows a geom to
remove incomplete observations without displaying a warning.

# Reference

Wickham, H., Çetinkaya-Rundel, M., and Grolemund, G. (2023). *R for Data
Science: Import, Tidy, Transform, Visualize, and Model Data* (2nd ed.).
O'Reilly Media. https://r4ds.hadley.nz/data-visualize.html
